---
title: "Introducing the ipfr package"
author: "Kyle Ward"
date: 2019-08-28
categories: ["R"]
tags: ["R Markdown", "ipfr", "packages"]
image:
  placement: 2
  caption: ""
  focal_point: ""
  preview_only: false
projects: ["ipfr"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
library(ipfr)
library(dplyr)
set.seed(42)
```

I'm happy to announce that the [ipfr package](https://github.com/dkyleward/ipfr)
is available on CRAN! The goal of this package is to make survey expansion,
matrix balancing, and population synthesis easier.

A basic use case is the task of balancing a matrix to row and column targets:

```{r, eval=FALSE}
library(ipfr)
library(dplyr)
```
```{r matrix}
mtx <- matrix(data = runif(9), nrow = 3, ncol = 3)
row_targets <- c(3, 4, 5)
column_targets <- c(5, 4, 3)
result <- ipu_matrix(mtx, row_targets, column_targets)

rowSums(result)
colSums(result)
```

The example below creates a simple survey and expands it to meet known
population targets. Each row in the `survey` data frame represents a household
and contains information on the number of household members (`size`) and number
of `autos`. The `targets` list contains population targets that the survey
expansion should match. For example, there should be a total of 75 households
with 1 person.

```{r survey}
survey <- tibble(
  size = c(1, 2, 1, 1),
  autos = c(0, 2, 2, 1),
  weight = 1
)
targets <- list()
targets$size <- tibble(
  `1` = 75,
  `2` = 25
)
targets$autos <- tibble(
  `0` = 25,
  `1` = 50,
  `2` = 25
)
result <- ipu(survey, targets)
```

The package also supports a number of advanced features:

  * Match to household- and person-level targets simultaneously
  * View and restrict the distribution of resulting weights
  * Control by geography
  * Handle target agreement and importance

Finally, the resulting weight table can be used to easily create a synthetic
population:

```{r}
synthesize(result$weight_tbl)
```