---
title: "Understanding Transit Ridership Trends"
author: "Kyle Ward"
date: 2019-09-03
categories: ["R"]
tags: ["R Markdown", "transit"]
image:
 placement: 2
 caption: ""
 focal_point: ""
 preview_only: false
projects: ""
editor_options: 
  chunk_output_type: console
draft: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, out.width = "100%")
library(tidyverse)
library(lubridate)
library(readxl)
library(here)
library(plotly)
library(RColorBrewer)
```

```{r}
xlsx <- here::here("static", "data", "June 2019 Adjusted Database.xlsx")
upt_tbl_raw <- read_excel(xlsx, sheet = "UPT")

# We will use this table as a starting point for several charts below
upt_tbl <- upt_tbl_raw %>%
  filter(Active == "Active") %>%
  gather(key = "month_year", value = "trips", JAN02:JUN19) %>%
  mutate(
    year = substr(
      month_year, start = str_length(month_year) - 1, stop = str_length(month_year)
    ),
    month = substr(month_year, start = 1, stop = 3),
    date = parse_date_time2(month_year, orders = "my")
  ) %>%
  arrange(Agency, date)

# Assign a color to each agency for consistent charts
color_pal <- colorRampPalette(brewer.pal(8, "Set2"))
color_tbl <- upt_tbl %>%
  filter(month_year == "JUN19") %>%
  group_by(Agency) %>%
  summarize(trips = sum(trips)) %>%
  arrange(desc(trips)) %>%
  mutate(color = color_pal(n())) %>%
  select(-trips)
upt_tbl <- upt_tbl %>%
  left_join(color_tbl, by = "Agency")
```

```{r}
# convenience functions

# shortens the agency name to the first 4 characters. Improves legend.
shorten_agency <- function(Agency){
  Agency = gsub("[^a-zA-Z0-9 _]", "", Agency)
  word_count = str_count(Agency, "\\W+") + 1
  agency_short = word(Agency, 1, pmin(word_count, 4))
  return(agency_short)
}

# Create a function that performs seasonal decomposition
extract_trend <- function(x, start_year) {
  myts <- ts(x, start = c(start_year, 1), end = c(2019, 6), frequency = 12)
  fit <- stl(myts, s.window="period")  
  df <- as.data.frame(fit$time.series)
  return(df$trend)
}
```

Before talking about national transit trends, it's important to grasp the size
of the NYC market. The chart below shows the top 10 transit agencies by monthly
ridership (June 2019 unlinked trips). NYC dwarfs the other transit markets, and
for this post, is set aside as an outlier.

```{r, out.width="100%", warning=FALSE}
data1 <- upt_tbl %>%
  filter(month_year == "JUN19") %>%
  group_by(Agency) %>%
  summarize(Trips = sum(trips, na.rm = TRUE)) %>%
  arrange(desc(Trips)) %>%
  head(10) %>%
  arrange(Trips) %>%
  mutate(Agency = shorten_agency(Agency))

plot_ly(
  data1, y = ~Agency, x = ~Trips, type = "bar", orientation = "h",
  text = data$Agency, textposition = "auto", hoverinfo = 'x'
) %>%
  layout(
    title = "Top 10 Transit Agencies (trips)",
    yaxis = list(
      categoryorder = "array",
      categoryarray = ~Agency,
      showticklabels = FALSE
    )
  )
```

So what's going on outside of NYC? Here is the raw data of transit ridership
over time by agency. We'll need to do some further processing before any
useful information can be extracted.

```{r, out.width="100%", warning=FALSE}
data2 <- upt_tbl %>%
  filter(Agency != "MTA New York City Transit") %>%
  group_by(Agency, date) %>%
  summarize(
    trips = sum(trips, na.rm = TRUE),
    color = first(color)
  ) %>%
  arrange(Agency, date) %>%
  ungroup() %>%
  mutate(Agency = shorten_agency(Agency))
  
plot_ly(
  data2, x = ~date, y = ~trips, type = "scatter", mode = "lines",
  color = ~Agency, colors = ~unique(color)
) %>%
  layout(
    showlegend = FALSE
  )
```

## Collapsing Agencies

A histogram of June 2019 ridership tells us that there are a few
agencies between 10 and 40 million monthly trips, several between 2 and 10
million, and then the large majority of agencies are below 2 million monthly
trips.

```{r}
data3 <- upt_tbl %>%
  filter(month_year == "JUN19", Agency != "MTA New York City Transit") %>%
  group_by(Agency) %>%
  summarize(Trips = sum(trips, na.rm = TRUE))

plot_ly(
  data3, x = ~Trips, type = "histogram", nbinsx = 30
)
```

To make the time series charts more legible, we'll collapse agencies under 2
million into a single agency and display their median ridership. We'll create
another median agency representing agencies between 2 and 10 million.

## Seasonality

Understanding transit ridership is also complicated by the seasonal nature of
it. For most agencies, ridership is low in December and January and peaks around
October. Removing these seasonal variations helps us understand the underlying
trends. The chart below shows the monthly ridership for the Chicago Transit
Authority. The actual ridership is shown by the jagged line while the trend is
highlighted in orange. The trend line makes it clear that (seaonally adjusted)
ridership peaked in April of 2012. As a point of reference, the vertical green
line marks when Uber began service in Chicago.

```{r}
data4 <- upt_tbl %>%
  filter(Agency == "Chicago Transit Authority") %>%
  group_by(Agency, date) %>%
  summarize(
    trips = sum(trips, na.rm = TRUE),
    color = first(color)
  ) %>%
  mutate(trend = extract_trend(trips, 2002))

plot_ly(data4) %>%
  add_trace(
    x = ~date, y = ~trips, type = "scatter", mode = "lines",
    alpha = .33, name = "seasonal", color = ~Agency, colors = ~unique(color)
  ) %>%
  add_trace(
    x = ~date, y = ~trend, type = "scatter", mode = "lines",
    name = "trend", alpha = 1
  ) %>%
  add_trace(
    x = c(ymd("2011-09-01"), ymd("2011-09-01")),
    y = c(0, 52000000),
    type = "scatter", mode = "lines", name = "Uber in CHI",
    hoverinfo = "text", hovertext = "Uber Starts",
    showlegend = FALSE
  ) %>%
  layout(
    title = "Chicago Transit Authority Ridership",
    yaxis = list(rangemode = "tozero"),
    annotations = list(
      x = ymd("2011-09-01"), y = 35000000, text = "Uber Starts",
      xref = "x", yref = "y",
      showarrow = TRUE, ax = 40, ay = 25
    )
  )
```

The chart below tells a fairly concise story about what is happening to transit
ridership. The medium markets like Chicago began seeing declines in ridership
between 2012 and 2014, and this trend has has continued. This is the same time
frame when Uber and Lyft began showing up in these cities.

```{r}
data5 <- upt_tbl %>%
  filter(
    Agency != "MTA New York City Transit",
    date >= ymd("2010-01-01")
  ) %>%
  group_by(Agency, date) %>%
  summarize(
    trips = sum(trips, na.rm = TRUE),
    color = first(color)
  ) %>%
  arrange(Agency, date) %>%
  mutate(current_ridership = last(trips)) %>%
  ungroup() %>%
  mutate(
    Agency = case_when(
      current_ridership <= 2000000 ~ "Under 2M",
      current_ridership <= 10000000 ~ "2M to 10M",
      TRUE ~ Agency
    )
  ) %>%
  group_by(Agency, date) %>%
  summarize(
    trips = median(trips),
    color = first(color)
  ) %>%
  mutate(trend = extract_trend(trips, 2010)) %>%
  ungroup()

plot_ly(data5) %>%
  add_trace(
    x = ~date, y = ~trend, type = "scatter", mode = "lines",
    color = ~Agency, colors = ~unique(color),
    text = ~Agency, hoverinfo = "text",
    showlegend = FALSE
  )
```

The small (<10 million) and tiny (<2 million) agencies look relatively stable in
the chart above, but that is mainly due to the scale of the graph. The charts
below zoom in to provide a more accurate picture. The median ridership in these
markets is also falling. The decline in the tiny markets is noticeably later.
One theory could be that Uber/Lyft simply arrived in these regions later.

```{r}
data6 <- data5 %>%
  filter(Agency %in% c("Under 2M", "2M to 10M"))

p1 <- plot_ly(data5 %>% filter(Agency == "2M to 10M")) %>%
  add_trace(
    x = ~date, y = ~trend, type = "scatter", mode = "lines",
    color = ~Agency, colors = ~unique(color),
    text = ~Agency, hoverinfo = "text"
  ) %>%
  layout(yaxis = list(rangemode = "tozero"))

p2 <- plot_ly(data5 %>% filter(Agency == "Under 2M")) %>%
  add_trace(
    x = ~date, y = ~trend, type = "scatter", mode = "lines",
    color = ~Agency, colors = ~unique(color),
    text = ~Agency, hoverinfo = "text"
  ) %>%
  layout(yaxis = list(rangemode = "tozero"))

subplot(p1, p2)
```

The Uber/Lyft connection seems obvious, but this post has not done the work
required to say anything for sure. In a future post, I hope to look at the link
in more detail. This goal of this post was primarily to provide insight into
what is happening with transit ridership.

## Reproducibility
Reproduce what you see above and take it further!