---
title: COVID-19 Trends and Insights
author: D. Kyle Ward
date: 'June 25, 2020'
slug: covid-19-trends
categories:
  - Health
tags: []
subtitle: ''
summary: ''
authors: []
lastmod: '2020-06-16T22:05:27-04:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
draft: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE, fig.height = 4, warning = FALSE, out.width = "100%"
)
library(tidyverse)
library(lubridate)
library(zoo)
library(plotly)
library(here)
library(tidycensus)
library(rvest)
library(knitr)
library(kableExtra)
library(readxl)
library(htmltools)
```

```{r}
# Move this css tag outside the chunk to control the width of text
# on the page.

# <style type="text/css">
# .main-container {
#   max-width: 1000px;
#   margin-left: auto;
#   margin-right: auto;
# }
# </style>
```

```{r, message=FALSE}
# Get data from covidtracking.com and store a copy in this repo. To update the
# data, delete the nation and state CSV files in this repo.

nation_csv <- here("static", "data", "2020-06-16-covid-19", "nation.csv")
if (!file.exists(nation_csv)) {
  nation <- read_csv(url("https://covidtracking.com/api/v1/us/daily.csv"))
  nation <- nation %>%
    filter(states == 56) %>%
    mutate(date = ymd(date)) %>%
    arrange(date)
  write_csv(nation, nation_csv)
}

states_csv <- here("static", "data", "2020-06-16-covid-19", "states.csv")
if (!file.exists(states_csv)) {
  states <- read_csv(url("https://covidtracking.com/api/v1/states/daily.csv"))
  states <- states %>%
    filter(!is.na(dateChecked)) %>%
    mutate(
      date = ymd(date),
      region = state.region[match(state, state.abb)]
    ) %>%
    arrange(state, date)
  
  # Grab census population data by state and add to state table
  api_key <- Sys.getenv("CENSUS_API_KEY")
  acs <- get_acs(
    "state",
    variables = "B01003_001E",
    year = 2018
  )
  states <- states %>%
    left_join(
      acs %>% select(GEOID, population = estimate),
      by = c("fips" = "GEOID")
    ) %>%
    filter(!is.na(population))
  
  # Add governor data to the state table
  state_abb <- setNames(state.abb, state.name)
  wikiurl <- "https://en.wikipedia.org/wiki/List_of_current_United_States_governors_by_age"
  data <- wikiurl %>%
    xml2::read_html()
  gov_tbl <- html_table(data)[[1]] %>%
    mutate(state = as.character(state_abb[State])) %>%
    select(state, state_full = State, Party)
  states <- states %>%
    left_join(gov_tbl, by = "state") %>%
    filter(!is.na(Party))
  
  
  # Add lock down dates
  balloturl <- "https://ballotpedia.org/Status_of_lockdown_and_stay-at-home_orders_in_response_to_the_coronavirus_(COVID-19)_pandemic,_2020"
  data <- balloturl %>%
    xml2::read_html()
  data <- html_table(data, fill = TRUE)[[3]]
  data <- data[2:nrow(data), 1:2]
  colnames(data) <- c("state", "date")
  data <- data %>%
    separate(date, c("start", "stop"), sep = "-") %>%
    separate(stop, c(NA, "month", "day", NA), sep = " ") %>%
    separate(day, c("day", NA), sep = "\\[") %>%
    unite(stop, month, day, sep = " ") %>%
    mutate(
      start = paste(start, "2020"),
      start = mdy(start),
      stop = paste(stop, "2020"),
      stop = ifelse(grepl("TBD", stop), "June 16 2020", stop),
      stop = mdy(stop),
      days = stop - start,
      days = ifelse(is.na(days), 0, days)
    ) %>%
    select(state, lockdown_days = days)
  states <- states %>%
    left_join(data, by = c("state_full" = "state"))
  
  write_csv(states, states_csv)
}

nation <- read_csv(nation_csv)
states <- read_csv(states_csv)
```

```{r colors}
# colors used for various charts
default_blue <- I(rgb(31, 119, 180, maxColorValue = 255))
neg_test_tan <- I(rgb(216, 179, 101, maxColorValue = 255))
hospital_red <- I(rgb(222, 45, 38, maxColorValue = 255))
death_gray <- I(rgb(99, 99, 99, maxColorValue = 255))
```


The motivations for this post are simple: 

  1. There is a wealth of data on the COVID-19 pandemic.
  2. That data is misrepresented on a daily basis.
  
The most egregious sin is presenting counts of new cases without accounting for
increased testing. This distorts comparisons between states (and countries) but
also misrepresents temporal trends within them. I present the numbers both ways
in the analysis below and also offer an interactive application for viewing the
data. In addition, I compare the experience of states in general and California
to New York explicitly to offer some data-driven insight as to why states had
such drastically different experiences.

The primary data source I use below is from
[https://covidtracking.com/](https://covidtracking.com/). All data/code is linked at the bottom of the article.

## National Data

### New Cases

The chart below shows the number of daily new cases in blue and the 15-day trend
in orange. This chart suggests that we saw a modest reduction in new cases and
are now poised to see an explosion. It's alarming and has prompted renewed
wall-to-wall news coverage of a "second wave". It looks (incorrectly) like an
approaching tsunami.

```{r}
temp <- nation %>%
  mutate(sma = rollmean(positiveIncrease, 15, na.pad = TRUE, align = "right"))

plot_ly(data = temp) %>%
  add_trace(
    x = ~date, 
    y = ~round(positiveIncrease, -2),
    type = "bar", name = "New Cases",
    color = default_blue,
    alpha = .15, alpha_stroke = .15
  ) %>%
  add_trace(
    x = ~date, y = ~sma, type = "scatter", mode = "lines", hoverinfo = "none",
    name = "15-day Trend", color = default_blue
  ) %>%
  layout(
    title = "Daily New Cases",
    legend = list(x = .7, y = .95),
    yaxis = list(title = "Cases"),
    xaxis = list(title = "")
  ) %>%
  config(displayModeBar = FALSE)
```

The above graph has been presented in various forms by practically every news
outlet. That's unfortunate, because it's misleading. The first chart below
presents the same data in blue but also includes the number of negative tests in
orange. The second chart translates both numbers into the rate of new cases per
100 tests performed.

```{r, fig.height=3}
plot_ly(data = nation) %>%
  add_trace(
    x = ~date,
    y = ~round(positiveIncrease, -2),
    type = "scatter", mode = "line",
    name = "New Cases"
  ) %>%
  add_trace(
    x = ~date,
    y = ~round(negativeIncrease, -2),
    type = "bar",
    name = "Negative Tests",
    color = neg_test_tan
  ) %>%
  layout(
    title = "Daily New Cases and Negative Test Results",
    yaxis = list(title = "Tests"),
    xaxis = list(title = NA)
  ) %>%
  layout(
    legend = list(x = .1, y = .95),
    hovermode = "x unified"
  ) %>%
  config(displayModeBar = FALSE)
```

```{r, fig.height=3}
temp <- nation %>%
  mutate(
    positive_rate = round(positiveIncrease / totalTestResultsIncrease * 100, 1),
    sma = rollmean(positive_rate, 7, na.pad = TRUE, align = "right")
  )
  
plot_ly(data = temp) %>%
  add_trace(
    x = ~date, y = ~positive_rate, type = "bar", name = "Positive",
    color = default_blue, alpha = .15, alpha_stroke = .15
  ) %>%
  add_trace(
    x = ~date, y = ~sma, type = "scatter", mode = "lines", hoverinfo = "none",
    name = "7-day Trend", color = default_blue
  ) %>%
  layout(
    title = "Daily New Cases per 100 Tests",
    yaxis = list(
      title = "New Cases per 100 Tests"
    ),
    xaxis = list(title = NA),
    legend = list(x = .7, y = .95)
  ) %>%
  config(displayModeBar = FALSE)
```

To be sure, the rate of new cases has begun rising in the past few days;
however, these charts present a much different picture of reality. It is also
the correct one. If five percent of the population is sick, doubling the number
of tests doubles the amount of positive cases detected, but the rate of
infection is unchanged. Presenting raw counts without accounting for this is
bad data science.

It is also worth noting that the high rates of positive cases in April were in
part the result of limited tests being reserved for people with COVID-like
symptoms. As testing increased, a more accurate measure of the infection rate
has emerged.

### Hospitalizations

Tracking daily hospital usage offers perhaps the best way to gauge the impact
of COVID-19. In the chart below, each bar represents the number of people
in the hospital on that day due to the virus. This metric fell from late April
to early June, but it has risen in the past four days.

```{r, warning=FALSE}
temp <- nation %>%
  mutate(
    sma = rollmean(hospitalizedCurrently, 3, na.pad = TRUE, align = "right")
  )

plot_ly(data = temp) %>%
  add_trace(
    x = ~date, y = ~hospitalizedCurrently,
    type = "bar",
    # type = "scatter", mode = "lines",
    name = "Hospitalized",
    color = hospital_red, alpha = .15, alpha_stroke = .15
  ) %>%
  add_trace(
    x = ~date, y = ~sma, type = "scatter", mode = "lines", hoverinfo = "none",
    name = "3-day Trend", color = hospital_red
  ) %>%
  layout(
    title = "Daily Hospitalized",
    yaxis = list(title = "Hospitalized"),
    xaxis = list(title = ""),
    legend = list(x = .7, y = .95)
  ) %>%
  config(displayModeBar = FALSE)
```

### Deaths

Similar to hospital usage, COVID-19 deaths offer a metric that is easier to
interpret without worrying about sample size issues (as with counts of new
cases).

In the chart below, the 15-day trend smooths out the cyclical pattern and makes
the decreasing trend clear. Deaths have fallen since late April; however, deaths
are a lagging indicator. In a second wave scenario, you would expect to see
hospitalizations go up first followed by rising deaths. The last bar on the
chart (June 25th), is an odd data point and likely a data reporting issue. (The
[New York Times data](https://github.com/nytimes/covid-19-data) lists a similar
figure of 2,466.)

```{r}
temp <- nation %>%
  mutate(sma = rollmean(deathIncrease, 15, na.pad = TRUE, align = "right"))

plot_ly(data = temp) %>%
  add_trace(
    x = ~date, y = ~deathIncrease, type = "bar", name = "Deaths",
    color = death_gray,
    alpha = .15, alpha_stroke = .15
  ) %>%
  add_trace(
    x = ~date, y = ~sma, type = "scatter", mode = "lines", 
    name = "15-day Trend", color = death_gray
  ) %>%
  layout(
    title = "Daily Deaths",
    showlegend = FALSE,
    xaxis = list(title = ""),
    yaxis = list(title = "Deaths")
  ) %>%
  config(displayModeBar = FALSE)
```

### National Summary

Perhaps the most important take away from the above review is that daily
hospitalization rates are the best statistic to broadcast. They don't require
adjustment based on the amount of testing or lag behind as much as daily deaths.
Most media stories I'm seeing are misrepresenting the data by focusing on the
absolute number of new cases. Without adjusting for the amount of testing, this
is the worst metric to use.

## State Data

While the nation as a whole is trending in the right direction, decisions to
re-open are made state by state. Fortunately, we can look at the same metrics
for each state. Additionally, the experiences of different states may create
natural experiments for us to examine. 

In addition to the analysis I present in this post, I've also created an
interactive viewer for the state data. Click the image below to open the app and
choose which metric to view, see trends for each state, and compare states to
each other.

```{r}
app_pic <- here(
  "static", "img", "2020-06-16-covid-19", "interactive_app.png"
)
```

<center>

<a href="http://us-covid-app.herokuapp.com/" target = "_blank"><img src=`r app_pic` width=400px"/></a>

</center>

### New Case Rate

The table below shows the rate of new cases per day for a handful of states.
Issues with the daily state data create some noise, but the different patterns
are easy to see.

```{r}
temp <- states %>%
  arrange(state, date) %>%
  filter(
    state %in% c("NC", "SC", "GA", "NY"),
    date > ymd("2020-03-20")
  ) %>%
  group_by(state) %>%
  mutate(
    positive_rate = round(positive / totalTestResults * 100, 1)
  ) %>%
  ungroup() %>%
  select(state, date, positive_rate)

plot_ly(data = temp) %>%
  add_trace(
    x = ~date, y = ~positive_rate, type = "scatter", mode = "markers",
    color = ~state
  ) %>%
  layout(
    title = "Daily Positive Testing Rate",
    yaxis = list(
      title = "Positives per 100"
    ),
    xaxis = list(
      title = NA
    )
  ) %>%
  config(displayModeBar = FALSE)
```

### Death Rate

At the state level, comparing absolute deaths would be misleading. States like
New York and California have more people, which means more deaths. As I did with
new cases, I convert deaths to a rate: deaths per one million people. Even with
this adjustment, the chart below shows how severe the problem became in NY
compared to other states.

```{r}
temp <- states %>%
  arrange(state, date) %>%
  filter(
    state %in% c("NC", "SC", "GA", "NY", "CA", "TX"),
    date > ymd("2020-03-20")
  ) %>%
  mutate(
    death_rate = deathIncrease / population * 1000000
  )

plot_ly(data = temp) %>%
  add_trace(
    x = ~date, y = ~death_rate, type = "scatter", mode = "markers",
    color = ~state
  ) %>%
  layout(
    title = "Daily Deaths by State",
    yaxis = list(
      title = "Deaths per Million"
    ),
    xaxis = list(
      title = NA
    )
  ) %>%
  config(displayModeBar = FALSE)
```

## Regional Differences

The chart below shows the rates of hospitalization and death on a weekly basis
colored by region of the country. Any state that experienced a weekly death rate
above 50 per million is labeled.

  * x-axis: deaths per million
  * y-axis: hospitalized per million
  * circle size: state population (Census ACS)
  * color: region

Press "Play" to see two primary patterns emerge:

  1. The separation of the northeastern states around New York City.
  2. Hospitalization and death rates are falling for nearly all states.

```{r}
# create the weekly stats data frame
weekly_stats_df <- states %>%
  arrange(state, date) %>%
  mutate(
    week = floor_date(date, "weeks"),
    week_id = week(date)
  ) %>%
  group_by(state, week) %>%
  summarize(
    days = n(),
    week_id = first(week_id),
    positiveIncrease = sum(positiveIncrease),
    totalTestResultsIncrease = sum(totalTestResultsIncrease),
    deathIncrease = sum(deathIncrease),
    hospitalizedCurrently = mean(hospitalizedCurrently, na.rm = TRUE),
    hospitalizedCurrently = ifelse(
      is.nan(hospitalizedCurrently), NA, hospitalizedCurrently),
    Party = first(Party),
    population = first(population),
    lockdown_days = first(lockdown_days),
    state_full = first(state_full),
    region = first(region)
  ) %>%
  # filter out early weeks and any week with <7 days
  filter(week_id > 12, days == 7) %>%
  mutate(
    week_id = week_id - 12,
    positive_rate = round(positiveIncrease / totalTestResultsIncrease * 100, 1),
    # some errors in the reported data lead to 1-2 negative numbers
    positive_rate = pmax(positive_rate, 0),
    death_rate = round(deathIncrease / population * 1000000, 4),
    hosp_rate = round(hospitalizedCurrently / population * 1000000, 4),
    lockdown_class = case_when(
      lockdown_days > 60 ~ "1.Long",
      lockdown_days > 30 ~ "2.Medium",
      lockdown_days == 0 ~ "4.None",
      TRUE ~ "3.Short"
    ),
    lockdown_class = factor(
      lockdown_class,
      levels = c("1.Long", "2.Medium", "3.Short", "4.None"),
      ordered = TRUE
    )
  ) %>%
  ungroup()
```


```{r}
assert_zero <- weekly_stats_df %>%
  mutate(
    hosp_rate = ifelse(is.na(hosp_rate), 0, hosp_rate)
  )

# Create a DF of just those states to be labeled. Label any state that
# exceeded 50 dpm in any week.
labels <- assert_zero %>%
  group_by(state) %>%
  mutate(exceeded_50dpm = ifelse(max(death_rate) > 50, 1, 0)) %>%
  filter(exceeded_50dpm == 1)

num_of_states_over_50dpm <- length(unique(labels$state ))

plot_ly(data = assert_zero) %>%
  add_trace(
    x = ~death_rate, y = ~hosp_rate, 
    type = "scatter", mode = "markers",
    size = ~population,
    # color = "orange",
    color = ~region,
    alpha = .05, alpha_stroke = .05, showlegend = FALSE,
    hoverinfo = "none"
  ) %>%
  add_trace(
    x = ~death_rate, y = ~hosp_rate, 
    type = "scatter", mode = "markers",
    marker = list(
      line = list(
        color = rgb(49, 52, 56, maxColorValue = 255)
      )
    ),
    color = ~region,
    ids = ~state, frame = ~week_id,
    size = ~population,
    hoverinfo = "text",
    text = ~paste0(
      "<br>", state, " week ", week,
      "<br>Deaths: ", round(death_rate, 0), "/M",
      "<br>Hopitalized: ", round(hosp_rate, 0), "/k"
    )
  ) %>%
  add_text(
    data = labels, text = ~state,
    x = ~death_rate, y = ~hosp_rate,
    ids = ~state, frame = ~week_id,
    textposition = "top", showlegend = FALSE
  ) %>%  
  layout(
    title = "Weekly Metrics by State",
    yaxis = list(
      title = "Hospitalized per Million"
    ),
    xaxis = list(
      title = "Deaths per Million",
      range = c(0, 273)
    )
  ) %>%
  config(displayModeBar = FALSE) %>%
  animation_opts(
    frame = 1500
  ) %>%
  animation_slider(
    currentvalue = list(
      prefix = "Weeks since 3/29: ",
      font = list(size = 12)
    )
  )
```

*Note: The states not reporting hospitalization data run along the x-axis
only*

## New York City

New York City is the [densest city in the
US](https://en.wikipedia.org/wiki/List_of_United_States_cities_by_population),
has the most [air
travel](https://en.wikipedia.org/wiki/List_of_busiest_city_airport_systems_by_passenger_traffic),
and has [transit ridership](/2019/09/03/understanding-transit-ridership-trends/)
that dwarfs other cities. These factors make it a hotbed for viral transmission.
The map below shows the peak death rate experienced in each state, and the
impact of New York City can be seen in the surrounding states. Outside of the
northeast, the peak death rates are much lower. Louisiana stands out in the
south, which the CDC
[attributes](https://www.cdc.gov/mmwr/volumes/69/wr/mm6915e4.htm?s_cid=mm6915e4_w)
to Mardi Gras.

```{r}
peak_deaths <- weekly_stats_df %>%
  group_by(state) %>%
  mutate(peak_death_rate = max(death_rate)) %>%
  filter(death_rate == peak_death_rate) %>%
  slice(1)

plot_geo() %>%
  add_trace(
    z = ~peak_deaths$peak_death_rate,
    hoverinfo = "text",
    # text = state.name,
    text = paste0(
      "<br>", state.name,
      "<br>Peak Rate: ", round(peak_deaths$peak_death_rate, 1)
    ),
    span = I(0),
    colors = "Greys",
    locations = peak_deaths$state, locationmode = 'USA-states'
  ) %>%
  layout(
    geo = list(
      scope = 'usa',
      projection = list(type = 'albers usa'),
      lakecolor = toRGB('white')    
    ),
    dragmode = FALSE
  ) %>%
  colorbar(
    title = paste0(
      "Peak Death Rate",
      "<br>(deaths per million)"
    )
  ) %>%
  config(displayModeBar = FALSE)
```

## Comparing to California

California is the most populous state, but it's experience with the virus so far
has been much less severe. Comparing California and New York reveals important
similarities and differences that could explain the different experience. It's
also helpful that they initiated stay-at-home orders [within a day of each
other](https://ballotpedia.org/Status_of_lockdown_and_stay-at-home_orders_in_response_to_the_coronavirus_(COVID-19)_pandemic,_2020).

### Density

New York City is the largest, densest city in the country. The table below shows
the top 10 US cities ranked by population density
([source](https://en.wikipedia.org/wiki/List_of_United_States_cities_by_population)).
San Francisco is second on the list, but has only 1/10th the population. In
fact, all the Californian cities in the table combined only hold ~1.4 million
people. This disparity in population and density is a significant reason why
NYC had a more severe outbreak.

```{r}
city_dens_url <- "https://en.wikipedia.org/wiki/List_of_United_States_cities_by_population"
data <- city_dens_url %>%
    xml2::read_html()
density_tbl <- html_table(data, fill = TRUE)[[5]]
density_tbl <- density_tbl[, c(2:4, 9)] %>%
  select(
    City, State = `State[c]`, Density = `2016 population density`,
    Population = `2019estimate`
  ) %>%
  mutate(
    City = gsub("\\[.*", "", City),
    density_num = gsub("/sq\\smi|,", "", Density),
    density_num = as.numeric(density_num)
  ) %>%
  arrange(desc(density_num))

density_tbl %>%
  mutate(
    State = cell_spec(State, "html", color = case_when(
      State == "California" ~ "red",
      State == "New York" ~ "blue",
      TRUE ~ "gray"
    )),
    City = cell_spec(City, "html", color = case_when(
      .$State == "California" ~ "red",
      .$State == "New York" ~ "blue",
      TRUE ~ "gray"
    ))
  ) %>%
  # filter(density_num >= 10000) %>%
  head(10) %>%
  select(-density_num) %>%
  kable(format = "html", escape = FALSE, row.names = FALSE) %>%
  kable_styling(full_width = FALSE)
```

### Air Travel

Air travel is another important consideration for viral spread. The table below
shows the top six US metro regions in terms of air travel
([source](https://en.wikipedia.org/wiki/List_of_busiest_city_airport_systems_by_passenger_traffic)).
While NYC tops the list, Los Angeles and San Francisco are not far behind. Being
on the east coast, NYC also attracts more European tourists, which is [where
their virus originated](https://www.nytimes.com/2020/04/08/science/new-york-coronavirus-cases-europe-genomes.html).

```{r}
air_url <- "https://en.wikipedia.org/wiki/List_of_busiest_city_airport_systems_by_passenger_traffic"
data <- air_url %>%
    xml2::read_html()
air_tbl <- html_table(data, fill = TRUE)[[5]] %>%
  filter(`Metropolitan area` %in% c(
    "New York City", "Los Angeles", "Atlanta", "Chicago",
    "Miami", "San Francisco Bay Area"
  )) %>%
  select(
    Metro = `Metropolitan area`,
    `Yearly Passengers` = Totalpassengers,
    `Airport(s)` = `Airport(s) included`
  )

air_tbl %>%
  mutate(
    Metro = cell_spec(Metro, "html", color = case_when(
      Metro %in% c("Los Angeles", "San Francisco Bay Area") ~ "red",
      Metro == "New York City" ~ "blue",
      TRUE ~ "gray"
    ))
  ) %>%
  kable(format = "html", escape = FALSE, row.names = FALSE) %>%
  kable_styling(full_width = FALSE)
```

### Transit

Crowded buses and subways are an excellent place for viruses to spread. In my
post on [transit ridership
trends](/2019/09/03/understanding-transit-ridership-trends/), I used the chart
below to show just how much larger the transit market is in NYC compared to the
rest of the country.

```{r}
xlsx <- here::here(
  "static", "data", "2020-06-16-covid-19", "April 2020 Adjusted Database.xlsx"
)
upt_tbl_raw <- read_excel(xlsx, sheet = "UPT")

upt_tbl <- upt_tbl_raw %>%
  filter(Active == "Active") %>%
  gather(key = "month_year", value = "trips", JAN02:APR20) %>%
  mutate(
    year = substr(
      month_year, start = str_length(month_year) - 1, stop = str_length(month_year)
    ),
    month = substr(month_year, start = 1, stop = 3),
    date = parse_date_time2(month_year, orders = "my")
  ) %>%
  arrange(Agency, date)

# shortens the agency name to the first 4 characters. Improves legend.
shorten_agency <- function(Agency){
  Agency = gsub("[^a-zA-Z0-9 _]", "", Agency)
  word_count = str_count(Agency, "\\W+") + 1
  agency_short = word(Agency, 1, pmin(word_count, 4))
  return(agency_short)
}
```

```{r, out.width="100%", warning=FALSE}
data1 <- upt_tbl %>%
  filter(month_year == "JUN19") %>%
  group_by(Agency) %>%
  summarize(Trips = round(sum(trips, na.rm = TRUE), -4)) %>%
  arrange(desc(Trips)) %>%
  head(10) %>%
  arrange(Trips) %>%
  mutate(Agency = shorten_agency(Agency))

plot_ly(
  data1, y = ~Agency, x = ~Trips, type = "bar", orientation = "h",
  text = ~Agency, textposition = "auto", hoverinfo = 'x'
) %>%
  layout(
    title = "Top 10 Transit Agencies (June 2019 trips)",
    yaxis = list(
      categoryorder = "array",
      categoryarray = ~Agency,
      showticklabels = FALSE
    )
  ) %>%
  config(displayModeBar = FALSE)

transit_multiple <- round(
  data1$Trips[data1$Agency == "MTA New York City"] / 
    data1$Trips[data1$Agency == "Los Angeles County Metropolitan"],
  1
)
```

Unlike the modest differences in air travel, NYC transit is 
`r transit_multiple` times larger than the largest market in California during
usual operation. People crammed into enclosed spaces on the NYC city subway were
primed to spread the disease.

As the pandemic spread, San Francisco [closed their
subway](https://sf.curbed.com/2020/3/26/21195566/muni-trains-covid-19-sf-underground-subway)
while Los Angeles cut back transit service after [steep drops in
ridership]((https://www.latimes.com/california/story/2020-04-17/coronavirus-cuts-los-angeles-metro-bus-train-service)).
NYC also saw reduced ridership and ran reduced service, but they only [closed
the
subway](https://abc7ny.com/subway-cleaning-plan-mta-coronavirus-nyc/6155622/)
between 1:00 am and 5:00 am,

The chart below updates my original 2019 chart with April 2020 data. NYC
still recorded more than twice the ridership of Los Angeles.

```{r, out.width="100%", warning=FALSE}
april20_data <- upt_tbl %>%
  filter(month_year == "APR20") %>%
  group_by(Agency) %>%
  summarize(Trips = round(sum(trips, na.rm = TRUE),-4)) %>%
  arrange(desc(Trips)) %>%
  head(10) %>%
  arrange(Trips) %>%
  mutate(Agency = shorten_agency(Agency))

plot_ly(
  april20_data, y = ~Agency, x = ~Trips, type = "bar", orientation = "h",
  text = ~Agency, textposition = "auto", hoverinfo = 'x'
) %>%
  layout(
    title = "Top 10 Transit Agencies (April 2020 trips)",
    yaxis = list(
      categoryorder = "array",
      categoryarray = ~Agency,
      showticklabels = FALSE
    )
  ) %>%
  config(displayModeBar = FALSE)
```

### Weather

Another difference between California and New York is the weather. The chart
below compares average temperatures in San Francisco and Los Angeles with New
York City ([source](https://www.ncdc.noaa.gov/cdo-web)). California remained
noticeably warmer than NYC through early April. There is still debate over the
influence of weather on COVID-19, but [some
researchers](https://www.npr.org/sections/goatsandsoda/2020/04/09/830297538/scientists-try-to-figure-out-if-summer-will-slow-the-spread-of-covid-19)
[think](https://www.cebm.net/covid-19/do-weather-conditions-influence-the-transmission-of-the-coronavirus-sars-cov-2/)
it could behave like it's well-known coronavirus cousins.

```{r}
weather_csv <- here(
  "static", "data", "2020-06-16-covid-19", "noaa_weather_2020.csv"
)
weather_tbl <- read_csv(
  weather_csv,
  col_types = cols(
    STATION = col_character(),
    NAME = col_character(),
    LATITUDE = col_double(),
    LONGITUDE = col_double(),
    ELEVATION = col_double(),
    DATE = col_character(),
    TAVG = col_double(),
    TMAX = col_double(),
    TMIN = col_double()
  )
) %>%
  separate(NAME, sep = ", ", into = c("NAME", "REGION")) %>%
  filter(!is.na(TAVG), REGION != "NJ US") %>%
  group_by(REGION, DATE) %>%
  summarize(
    LATITUDE = first(LATITUDE),
    LONGITUDE = first(LONGITUDE),
    TAVG = round(mean(TAVG), 1),
    TMAX = round(mean(TMAX), 1),
    TMIN = round(mean(TMIN), 1)
  ) %>%
  mutate(
    sma = rollmean(TAVG, 10, aligh = "right", na.pad = TRUE)
  )

plot_ly() %>%
  add_trace(
    data = weather_tbl, x = ~DATE, y = ~TAVG,
    color = ~REGION,
    type = "scatter", mode = "lines",
    alpha = .15, alpha_stroke = .15, showlegend = FALSE,
    hoverinfo = "none"
  ) %>%
  add_trace(
    data = weather_tbl, x = ~DATE, y = ~sma,
    color = ~REGION,
    type = "scatter", mode = "lines"
  ) %>%
  layout(
    title = "Average Temperature",
    yaxis = list(title = "Average Temperature"),
    xaxis = list(title = "Date"),
    hovermode = "x unified",
    legend = list(x = .1, y = .95)
  ) %>%
  config(displayModeBar = FALSE)

```

## What about Lockdowns?

Lockdowns are the first factor many consider when thinking about
differences between states. Intuition might suggest that states with weak or no
lockdown restrictions would have higher rates of death and positive tests.
Instead, the comparison of California and New York show that other factors
are more likely to impact severity.

The graphic below again shows hospitalization and death rates for all 50 states, but this
time, they are color coded by lockdown duration ([source](https://ballotpedia.org/Status_of_lockdown_and_stay-at-home_orders_in_response_to_the_coronavirus_(COVID-19)_pandemic,_2020)). Northeastern states
implemented longer lockdowns in response to a much more severe viral outbreak.
Outside the northeast, states adopted different approaches and no clear pattern
has emerged showing that longer/shorter durations made any meaningful difference.
  
```{r}
plot_ly(data = assert_zero) %>%
  add_trace(
    x = ~death_rate, y = ~hosp_rate, 
    type = "scatter", mode = "markers",
    color = ~lockdown_class,
    colors = c("blue", "gray", "orange", "red"),    
    size = ~population,
    alpha = .05, alpha_stroke = .05, showlegend = FALSE,
    hoverinfo = "none"
  ) %>%
  add_trace(
    x = ~death_rate, y = ~hosp_rate, 
    type = "scatter", mode = "markers",
    marker = list(
      line = list(
        color = rgb(49, 52, 56, maxColorValue = 255)
      )
    ),
    color = ~lockdown_class, 
    colors = c("blue", "gray", "orange", "red"),
    ids = ~state, frame = ~week_id,
    size = ~population,
    hoverinfo = "text",
    text = ~paste0(
      "<br>", state, " week ", week,
      "<br>DPM: ", round(death_rate, 0),
      "<br>HPM: ", round(hosp_rate, 0),
      "<br>Lockdown: ", lockdown_days, " days"
    )
  ) %>%
  add_text(
    data = labels, text = ~state,
    x = ~death_rate, y = ~hosp_rate, 
    ids = ~state, frame = ~week_id,
    textposition = "top", showlegend = FALSE
  ) %>%  
  layout(
    title = "Weekly Metrics by State",
    yaxis = list(title = "Hospitalized per Million"),
    xaxis = list(
      title = "Deaths per Million",
      range = c(0, 273)
    ),
    legend = list(
      title = list(text = "Lockdown<br>Duration")
    )
  ) %>%
  config(displayModeBar = FALSE) %>%
  animation_opts(frame = 1500) %>%
  animation_slider(
    currentvalue = list(
      prefix = "Weeks since 3/29: ",
      font = list(size = 12)
    )
  )
```

# Conclusion

Raw counts must be handled appropriately before making comparisons across time
or geography. Further, the severity of the virus has varied across the country
driven by underlying differences like density. Rural states with and without
lockdowns had similarly mild experiences. Finally, a second wave does appear to
be forming, which should be expected as states re-open. This fact alone does not
make it clear what the right policy response should be. Trade offs between viral
spread and mass unemployment are difficult to balance.

## Reproducibility and Data

All the data and code used to perform this analysis is available on GitHub:

[Data](https://github.com/dkyleward/blog/tree/master/static/data/2020-06-16-covid-19)  
[Code](https://github.com/dkyleward/blog/blob/master/content/post/2020-06-16-covid-19/2020-06-16-covid-19.en.Rmd) 

Other sources of interest:  
[https://covidtracking.com/](https://covidtracking.com/)  
[NY Times Data](https://github.com/nytimes/covid-19-data)